\chapter*{Acknowledgements}

First and foremost, I express my deepest gratitude to \textbf{Allah Almighty} for blessing me with the strength, perseverance, and guidance to complete this research work.

I am profoundly grateful to my thesis supervisor, \textbf{Dr. Malik Jahan Khan}, for their invaluable guidance, continuous support, and patience throughout this research journey. Their expertise in [field] and insightful feedback have been instrumental in shaping this work. Without their mentorship, this thesis would not have been possible.

I extend my sincere appreciation to the faculty and staff of the \textbf{University of Veterinary and Animal Sciences (UVAS), Lahore}, for providing access to the livestock disease data that forms the foundation of this research. Special thanks to the veterinary clinicians who contributed their expertise in data annotation and validation.

I am thankful to the members of my thesis committee for their constructive criticism and valuable suggestions that helped improve the quality of this research.

My heartfelt thanks go to my family---my parents, siblings, and loved ones---for their unconditional love, encouragement, and sacrifices. Their belief in me has been my greatest source of motivation.

I would also like to acknowledge my colleagues and friends who provided moral support and engaged in stimulating discussions that enriched my understanding of the research domain.

Finally, I am grateful to all the researchers whose prior work laid the foundation for this study. Their contributions to the fields of deep learning, natural language processing, and veterinary informatics have been invaluable.

\vspace{1cm}
\begin{flushright}
\textbf{Syed Muhammad Mujtaba}\\
Lahore, Pakistan\\
2025
\end{flushright}
